---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("./Data/Processed/NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

```

## GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 

### Geoms

Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

### Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

### Plotting continuous variables over time: Scatterplot
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()

O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)

# Fix this code
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "blue") #put color outside aesthetic to make the points blue
print(O3plot2) #need to say 'print' if you want the plot to show up in your pdf when you name the ggplot graph

# Add additional variables
PMplot <- 
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot) #month is the x-axis, pm2.5 is y-axis, change the shape and plot year data as a factor, color based on site.name (can put color in aes because we are using information from somewhere in the dataframe)

# Separate plot with facets
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3) #seperate data into variable of site name and organize into 3 rows; if rows weren't specified, it would automatically do 3 columns and 1 row; can put 2 variables into facet_wrap
print(PMplot.faceted)

PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ year) #make a grid of Site.Name by year (~ means 'by')
print(PMplot.faceted2)

# Filter dataset within plot building
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line() #connects each point every time we have an x-axis - won't produce an overall trendw
print(O3plot.Blackstone) #want to filter dataset within the plot; take a subset of the dataframe, and choose site.name == "Blackstone"

# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

# 1. 
# Plot surface temperatures by day of  year. 
# Color your points by year, and facet by lake in two rows.

ggplot(subset(PeterPaul.chem.nutrients, depth = 0), aes(x = daynum, y = temperature_c, color = year4) +
  geom_point() +
  facet_wrap(vars(lakename), nrow = 2)

  
#2. 
# Plot temperature by date. Color your points by depth.
# Change the size of your point to 0.5
ggplot(PeterPaul.chem.nutrients, aes(x = sampledate, y = temperature_C, color = depth)) +
  geom_point(size = 0.5)

```
### Plotting the relationship between two continuous variables: Scatterplot

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20) #limiting x and y axis to zoom in on a certain point in the graph from the above scatterplot
print(lightvsDOfixed)

# Depth in the fields of limnology and oceanography is on a reverse scale
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  #ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse() #can use to reverse the scales
print(tempvsdepth)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) + #color by depth; some points are grey on the graph which means there were no depth points, but there was a phosphorous or nitrogen measurements
  geom_point() +
  geom_smooth(method = lm) +
  geom_abline(aes(slope = 16, intercept = 0), lty = 2)
print(NvsP)


#geom_smooth - can give you a line on your data; method=lm means linear model and will give you a smooth line; also gives you confidence intervals - creates a band around linear line saying we are 95% confident this is the linear relationship

#geom_abline - lets us draw a line with a slope and intercept that we define; automatically shows up as a black solid line; lty = 2 will give us a dotted line

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
ggplot(EPAair, aes(x = PM2.5, y = Ozone, color = Site.Name)) +
geom_point() +
  geom_smooth(method = lm) #draws on aesthetics, so need to have aesthetics in ggplot; if you put it in geom_point, then you can put them in geom_smooth for it to use those aesthetics

# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%
  summarise(sd = sd(concentration), 
            mean = mean(concentration)) #group by both lake name and nutrient; if you want it to group by multiple factors, then need to be in the same group_by

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill instead of color? fill will put the color as the bar; color will only define what the boarder color is, but not the color of the bar
       position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here?
print(Nutrientplot)

#position dodge = seperating out the bars of the barplot instead of stacked on top of each other; stat = summary means 

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + #
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do?
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars? it needs aesthetics as any other layer would; wants the min and max value in this case
                 position = "dodge")
print(Nutrientplot2)

#barchart requires 'stat' to be defined; first bar graph we wanted it to produce a summary statistic and calculate the mean; for nutrient data set we already have the mean, so we just need it to take the identity for the cell it's pointing at

# Are there more effective ways to produce summary stats for categories?
#yes, there are better ways to plot this data

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)

#plotting every single value for each variable; wide spots is where we get the most measurements at that concentration; plots the density of the measurements that were made

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #color based on nutrients
print(Nutrientplot5)

#plotting densities, but as a smooth line instead of as dots

# Frequency polygons
# Using a tidy dataset (wide dataset)
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)

#plotted frequency plots with tidy dataset and then with gathered dataset; first one has five lines for the five frequency polygons; specify colors so that it doesn't post 5 black lines

# Using a gathered dataset (skinny dataset)
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 

EPAair_summary <- EPAair %>%
  group_by(Site.Name, year) %>%
  summarise(sd = sd(PM2.5, na.rm = TRUE), 
            mean = mean(PM2.5, na.rm = TRUE))


            
ggplot(EPAair_summary, aes(x = as.factor(year), y = mean, fill = Site.Name)) + 
  geom_bar(position = "dodge", stat = "identity") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), position = "dodge")

# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
