---
title: "10: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Perform advanced edits on ggplot objects to follow best practices for data visualization

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
getwd()
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("./Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
EPAair <- read.csv("./Data/Processed/EPAair_O3PM25_3sites1718_processed.csv")

EPAair$Date <- as.Date(EPAair$Date, format = "%Y-%m-%d")
PeterPaul.chem.nutrients$sampledate <- as.Date(PeterPaul.chem.nutrients$sampledate, format = "%Y-%m-%d")

```


### Themes
Often, we will want to change multiple visual aspects of a plot. Ggplot comes with pre-built themes that will adjust components of plots if you call that theme. 
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) 
print(O3plot)

O3plot1 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_gray() #default theme
print(O3plot1)

O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_bw() #has a box around the plot
print(O3plot2)

O3plot3 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_light() #no black lines; just grey lines
print(O3plot3)

O3plot4 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  theme_classic() #axis labesls still grey and lines and text is in black
print(O3plot4)
```

Notice that some aspects of your graph have not been adjusted, including: 

* text size
* axis label colors
* legend position and justification

If you would like to set a common theme across all plots in your analysis session, you may define a theme and call up that theme for each graph. This eliminates the need to add multiple lines of code in each plot. 

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
mytheme <- theme_classic(base_size = 14) + #increase font size to 14
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top") #alternative: legend.position + legend.justification to include the coordinates; legend.position = c(1,1), legend.justification = c(1,1)

# options: call the theme in each plot or set the theme at the start. 

O3plot5 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone)) +
  mytheme #added as a layer in ggplot
print(O3plot5)  

theme_set(mytheme) #theme_set - tells R to set that theme and use it throughout the remainder of the session; can do this in the beginning and not have to include in the ggplot code; you can override the theme if you have already set it by putting the theme information in the ggplot code

O3plot6 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot6)  

```

### Adjusting multiple components of your plots

While the theme allows us to set multiple aspects of plots, ggplot allows us to adjust other parts of plots outside of the theme. 

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
O3plot7 <- ggplot(EPAair, aes(x = Date, y = Ozone)) +
  geom_rect(xmin = as.Date("2016-01-01"), xmax = as.Date("2020-01-01"), 
            ymin = 0, ymax = 50, fill = "green") +  #fill in the rectangles according to the specifications and with the said color; generally put geom_rect before points so that lighter points will be visible
  geom_rect(xmin = as.Date("2016-01-01"), xmax = as.Date("2020-01-01"), 
            ymin = 50, ymax = 100, fill = "yellow") +
  geom_point() +
  geom_text(x = as.Date("2019-01-01"), y = 45, label = "good", hjust = 1, fontface = "bold") + #include text at y = 45 to say "good"; text should be bold and should horizontally justification to 1; default is to put text in centered
  geom_text(x = as.Date("2019-01-01"), y = 95, label = "moderate", hjust = 1, fontface = "bold") +
  scale_x_date(limits = as.Date(c("2017-01-01", "2018-12-31")), #limiting the x axis dates and abbreviate the label and include months in 2 month increments instead of 6 month incrememnts which is default
    date_breaks = "2 months", date_labels = "%b %y") + #break dates into two mont increments and label them with abbreviated months and years
  ylab(expression("O"[3]* " AQI Value")) + #include the ylabel; [3] gives subscripts; * = no space between labels
  theme(axis.text.x = element_text(angle = 45,  hjust = 1)) #use the theme to make the x-axis text on a 45 degree angle with a horizontal justification of 1 (hjust = 1, moves things to the left)
print(O3plot7)  

```

### Color palettes
There are several color palettes that are designed to be more effective than palettes in base R. These include Viridis (https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) and Color Brewer (http://colorbrewer2.org/). A few rules for choosing colors: 

* Consider if your plot needs to be viewed in black and white. If so, choose a sequential palette with varying color intensity.
* Choose a palette that is color-blind friendly
* Maximize contrast (e.g., no pale colors on a white background)
* Diverging color palettes should be used for diverging values (e.g., warm-to-cool works well for values on a scale encompassing negative and positive values)

Perception is key! Choose palettes that are visually pleasing and will communicate what you are hoping your audience to perceive. Hint: base R palettes are not ideal.

```{r, warning = FALSE}
#install.packages("viridis")
#install.packages("RColorBrewer")
#install.packages("colormap")
library(viridis)
library(RColorBrewer)
library(colormap)

scales::show_col(colormap(colormap = colormaps$viridis, nshades = 16)) #looking at Viridis color palette with 16 different shades; viridis color palette ends in ff
scales::show_col(colormap(colormap = colormaps$inferno, nshades = 16))
scales::show_col(colormap(colormap = colormaps$magma, nshades = 16))
display.brewer.all(n = 9)

NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth, shape = lakename)) +
  geom_point() 
print(NvsP)

# let's first make the plot look better.
# change your axis labels to reflect TN and TP in micrograms per liter.
# change your legend labels
NvsP2 <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, fill = depth, shape = lakename)) +
  geom_point(alpha = 0.8, size = 3) +
  #xlab("Total P (\U003BCg/L)") + 
  #ylab(expression("Total N (\U003BCg/L)")) +
  # change your legend labels here
    #scale_fill_manual(values = c(21, 23), #defining which colors to use in the graphs
                     #name="Legend",
                     #breaks=c(21, 23),
                     #labels=c("Depth, m", "Lake Name"))
  labs(x = "Total P, micro g/L", y = "Total N, micro g/L", fill = "Depth, m", shape = "Lake") +
  scale_shape_manual(values = c(21, 23), #for shapes 21 - 25, I can define the fill and the outline 
                     name = c("Lake Name", "Depth")) +
  #scale_color_distiller(palette = "Blues", direction = 1) + # use scale_color_brewer for discrete variables
  #scale_fill_distiller(palette = "Blues", direction = 1)
  scale_fill_viridis(option = "magma", direction = -1) +
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), legend.title = element_text(size = 12), labs(title))
print(NvsP2)

# change your y axis label to list concentration in micrograms per liter
# remove your x axis label
# change labels for nutrients in the legend
Nutrientplot <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration, color = nutrient)) + #makes color different with each nutrient category
  geom_boxplot() +
# place your additional edits here
 ylab(expression(Concentration~(mu*g/L))) +
  xlab(NULL) + 
  scale_y_continuous(expand = c(0, 0)) +
  #scale_color_brewer(palette = "YlGnBu") +
  #scale_color_manual(values = c("#7fcdbb", "#41b6c4", "#1d91c0", "#225ea8", "#0c2c84"), #defining which colors to use in the graphs, discrete color choices
                     #name="Nutrient",
                     #breaks=c("nh34", "no23", "po4", "tn_ug", "tp_ug"),
                     #labels=c("Ammonium", "Nitrate", "Phosphorous", "Total N (\U003BCg/L)", "Total P expression(TP~(mu*g/L)")) + 
  scale_color_viridis(discrete = TRUE,  name="Nutrient",
                     breaks=c("nh34", "no23", "po4", "tn_ug", "tp_ug"), #if you want this information, make sure it's in the color layer
                     labels=c("Ammonium", "Nitrate", "Phosphorous", "Total N", "Total P")) #viridis is always continuous color, but if youw want it to be discrete, put discrete = TRUE

print(Nutrientplot)

  #theme(legend.position = "right", title = "Lake Name")
```

### Adjusting facets

```{r, fig.height = 4, fig.width = 6}
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name ~ year) + 
  scale_x_continuous(breaks = c(1:12)) +
  #theme(strip.background = element_rect(fill = "black"), strip.text = element_text(color = "white"))
  ylab(expression("PM 2.5 AQI Value")) 
print(PMplot.faceted)

```

### Multiple plots on a page
In situations where facets don't fill our needs to place multiple plots on a page, we can use the package `gridExtra` to arrange plots. The `grid.arrange` function is extremely flexible in its ability to arrange plots in specific configurations. A useful guide can be found here: https://cran.r-project.org/web/packages/egg/vignettes/Ecosystem.html.

```{r, fig.height = 7, fig.width = 6}
#install.packages("gridExtra")
library(gridExtra)
grid.arrange(NvsP2, Nutrientplot)
grid.arrange(O3plot7, PMplot.faceted)

```

### Saving plots
The `ggsave` function allows you to save plots in jpg, png, eps, pdf, tiff, and other formats. The following information can be supplied: 

* filename, with file extension and in quotes (required)
* plot object (required)
* path, with file name 
* width, height, units 
* resolution (dpi)

For example: 
`ggsave("PMplot.jpg", PMplot.faceted,  path = "./Output/PMplot.jpg", height = 4, width = 6, units = "in", dpi = 300)`

