---
title: 'Assignment 4: Data Wrangling'
author: "Caroline Watson"
geometry: margin=2.54cm
output:
  pdf_document: default
  word_document: default
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on data wrangling. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
You will need to have the correct software installed to do this (see Software Installation Guide)
Press the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Assignments folder.
8. After Knitting, please submit the completed exercise (PDF file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Salk_A04_DataWrangling.pdf") prior to submission.

The completed exercise is due on Thursday, 7 February, 2019 before class begins.

## Set up your session 
1. Check your working directory, load the `tidyverse` package, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Generate a few lines of code to get to know your datasets (basic data summaries, etc.).

```{r}
#1
#viewing working dirctory
getwd()

#loading tidyvere
suppressMessages(library(tidyverse))
library(knitr)
library(kableExtra)

#uploading raw data files for ozone and PM 2.5
EPA_PM25_2017 <- read.csv("./Data/Raw/EPAair_PM25_NC2017_raw.csv")
EPA_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv")
EPA_O3_2017 <- read.csv("./Data/Raw/EPAair_O3_NC2017_raw.csv")
EPA_O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv")

#2

#summary of dataset for PM2.5 in 2017
head(EPA_PM25_2017)
summary(EPA_PM25_2017)
colnames(EPA_PM25_2017)
dim(EPA_PM25_2017)

#summary of dataset for PM2.5 in 2018
head(EPA_PM25_2018)
summary(EPA_PM25_2018)
colnames(EPA_PM25_2018)
dim(EPA_PM25_2018)

#summary of dataset for O3 in 2017
head(EPA_O3_2017)
summary(EPA_O3_2017)
colnames(EPA_O3_2017)
dim(EPA_O3_2017)

#summary of dataset for O3 in 2018
head(EPA_O3_2018)
summary(EPA_O3_2018)
colnames(EPA_O3_2018)
dim(EPA_O3_2018)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder.

```{r}
#3
#changing the date to date for all 4 data files
EPA_O3_2017$Date <- as.Date(EPA_O3_2017$Date, format = "%m/%d/%y")
EPA_O3_2018$Date <- as.Date(EPA_O3_2018$Date, format = "%m/%d/%y")

EPA_PM25_2017$Date <- as.Date(EPA_PM25_2017$Date, format = "%m/%d/%y")
EPA_PM25_2018$Date <- as.Date(EPA_PM25_2018$Date, format = "%m/%d/%y")

#4
#selecting Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
EPA_O3_2017 <- select(EPA_O3_2017, Date, DAILY_AQI_VALUE, 
      Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
      SITE_LONGITUDE)

#selecting columns for EPA ozone data for 2018
EPA_O3_2018 <- select(EPA_O3_2018, Date, DAILY_AQI_VALUE, 
    Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
    SITE_LONGITUDE)

#selecting columns for EPA PM 2.5 2017
EPA_PM25_2017 <- select(EPA_PM25_2017, Date, DAILY_AQI_VALUE, 
    Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
    SITE_LONGITUDE)

#selecting columns for EPA PM 2.5 2018
EPA_PM25_2018 <- select(EPA_PM25_2018, Date, DAILY_AQI_VALUE, 
    Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, 
    SITE_LONGITUDE)

#5 Fill all cells in AQS_PARAMETER_DESC with "PM2.5"
EPA_PM25_2017$AQS_PARAMETER_DESC <- c("PM2.5")

EPA_PM25_2018$AQS_PARAMETER_DESC <- c("PM2.5")

#6 write.csv for all four files
write.csv(EPA_O3_2017, row.names = FALSE, file = "./Data/Processed/EPA_O3_2017_Processed.csv")

write.csv(EPA_O3_2018, row.names = FALSE, file = "./Data/Processed/EPA_O3_2018_Processed.csv")

write.csv(EPA_PM25_2017, row.names = FALSE, file = "./Data/Processed/EPA_PM25_2017_Processed.csv")

write.csv(EPA_PM25_2018, row.names = FALSE, file = "./Data/Processed/EPA_PM25_2018_Processed.csv")
```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.  #combine into one dataset, or combine EPA O3 datasets and EPA pm25 datasets?
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Sites: Blackstone, Bryson City, Triple Oak
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `separate` function or `lubridate` package)

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7
#reading in processed files
EPA_O3_2017_processed <- read.csv("./Data/Processed/EPA_O3_2017_Processed.csv")
EPA_O3_2018_processed <- read.csv("./Data/Processed/EPA_O3_2018_Processed.csv")
EPA_PM25_2017_processed <- read.csv("./Data/Processed/EPA_PM25_2017_Processed.csv")
EPA_PM25_2018_processed <- read.csv("./Data/Processed/EPA_PM25_2018_Processed.csv")

#checking column names of all four datasets
colnames(EPA_O3_2017_processed)
colnames(EPA_O3_2018_processed)
colnames(EPA_PM25_2017_processed)
colnames(EPA_PM25_2018_processed)

#all column names are identical, so combining all datasets using rbind
EPA_Air_data_combined <- rbind(EPA_O3_2017_processed, 
  EPA_O3_2018_processed, EPA_PM25_2017_processed, 
  EPA_PM25_2018_processed)

#install.packages(lubridate)
library(lubridate)

#8 Wrangle dataset with pipe functions 
EPA_Air_summaries <- EPA_Air_data_combined %>%
  filter(Site.Name == "Blackstone" | Site.Name == "Bryson City" | Site.Name == "Triple Oak") %>%
    mutate_at(vars(Date), funs(month, year)) 

#9 spread datasets so that AQI values for ozone and PM2.5 are in seperate columns
EPA_Air_summaries_spread <- spread(EPA_Air_summaries, AQS_PARAMETER_DESC, DAILY_AQI_VALUE)

#10 call up the dimensions of your new tidy dataset
dim(EPA_Air_summaries_spread)

#11 saving processed data set with filename:"EPAair_O3_PM25_NC1718_Processed.csv"
write.csv(EPA_Air_summaries_spread, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1718_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate two new data frames: 
a. A summary table of mean AQI values for O3 and PM2.5 by month
b. A summary table of the mean, minimum, and maximum AQI of O3 and PM2.5 for each site (AQI values, not concentration values)
13. Display the data frames.

```{r}
#12a
#using split-apply-comine to generate summary table of mean AQI values for O3 and PM2.5 by month
EPA_Air_O3_PM25_mean_AQI <- EPA_Air_summaries_spread %>%
  group_by(month) %>%
  filter(!is.na(Ozone) & !is.na(PM2.5)) %>%
  summarise(mean_O3 = mean(Ozone), 
            mean_PM25 = mean(PM2.5))

EPA_Air_O3_PM25_mean_AQI2 <- EPA_Air_summaries_spread %>%
  group_by(month) %>%
  filter(!is.na(Ozone) | !is.na(PM2.5)) %>%
  summarise(mean_O3 = mean(Ozone), 
            mean_PM25 = mean(PM2.5))

View(EPA_Air_O3_PM25_mean_AQI2)

#12b

EPA_Air_O3_PM25_summary <- EPA_Air_summaries_spread %>%
  group_by(Site.Name) %>%
  filter(!is.na(Ozone) & !is.na(PM2.5)) %>%
  summarise(mean_O3 = mean(Ozone), 
            mean_PM2.5 = mean(PM2.5), 
            min_O3 = min(Ozone),
            min_PM2.5 = min(PM2.5), 
            max_O3 = max(Ozone), 
            max_PM2.5 = max(PM2.5))

EPA_Air_O3_PM25_summary2 <- EPA_Air_summaries_spread %>%
  group_by(Site.Name) %>%
  filter(!is.na(Ozone) & !is.na(PM2.5)) %>%
  summarise(mean_O3 = mean(Ozone), 
            mean_PM2.5 = mean(PM2.5), 
            min_O3 = min(Ozone),
            min_PM2.5 = min(PM2.5), 
            max_O3 = max(Ozone), 
            max_PM2.5 = max(PM2.5))

#13 
#display the data frames
kable(EPA_Air_O3_PM25_mean_AQI) %>%
  kable_styling()
kable(EPA_Air_O3_PM25_summary) %>%
  kable_styling()

kable(EPA_Air_O3_PM25_mean_AQI2) %>%
  kable_styling()
kable(EPA_Air_O3_PM25_summary2) %>%
  kable_styling()

```

